Implicit yaddayadda

### 2011-10-11 papers/PLDI2012/implicitization.tex
ajout d'une remarque dans disjunctive invariants
def Kleene

\section{Bases}
\subsection{Static analysis by abstract interpretation}
Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is often referred to as the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

@@

@@

\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} proposes to
distinguish the different paths in the program in order to avoid loss of
precision due to merge operations. Since the number of paths may be exponential,
the technique keeps them implicit and computes them when needed using
SMT-solving. Then, the iterations are computed over a reduced multigraph instead
of the classical transition graph.

>>>>

\section{Bases}
\subsection{Static analysis by abstract interpretation}
\label{sec:static_analysis}
Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is often referred to as the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

@@

Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g. $x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
\begin{equation}
\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
x' \in \gamma \circ \abstr{\tau}(\abstr{x})
\label{eqn:transformer_correct}
\end{equation}
It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the current program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.

@@

\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focused} technique
distinguishes the different paths in the program in order to avoid loss of
precision due to merge operations. Since the number of paths may be exponential,
the technique keeps them implicit and computes them when needed using
SMT-solving.
The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.
